---
title: "Afforestation planning in Iceland"
author: "J.P.Borchard"
date: "2025-10-16"
output: html_document
---

### Predicting ecologically suitable areas

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include = F}

library(raster)
library(terra)
library(sf)
library(dplyr)
library(leaflet)
library(whitebox)
library(RSAGA)
library(randomForest)
library(foreign)
library(blockCV)
library(caret)
library(spdep)

```

```{r include = F}
whitebox::wbt_init()


```

```{r include = F}
# --- 1) Load Iceland shapefile ---
iceland_shape <- vect("data/Iceland_shapefile/is_1km.shp")

```

## Sampling psuedo-presences for SDM from habitat map

The idea here is to read in mean NDVI for the 2023 growing period in Iceland. This was data aquired from google earth enginge. 
Then to pseudo-sample tree presences, 


```{r include = F}

ndvi <- rast("data/GEE/NDVI_2023_growing_period/NDVI_2023_growing_period_mosaic.tif")
skoglendi <- project(vect("data/forests/raektad_skoglendi_Island_07052025.shp"), ndvi)
natturu_birki <- project(vect("data/forests/Natturulegt_birkilendi_Island_15052025.shp"), ndvi)

presence_locations  <- spatSample(natturu_birki, size = 5000, method = "random")
```

```{r include = F, eval = F}

# Combine polygons quickly
#  forests_combined <- rbind(skoglendi, natturu_birki)

# Crop + mask NDVI
# ndvi_crop <- crop(ndvi, forests_combined)
# ndvi_forests <- mask(ndvi_crop, forests_combined)
# ndvi_forests[ndvi_forests < 0] <- NA

# pts_ndvi <- spatSample(ndvi_forests, 
 #                       size = 10000,        # number of points
#                        method = "random",  # random sampling
 #                       na.rm = TRUE,       # ignore NA pixels
#                        as.points = TRUE)   # return as SpatVector points




pts_skoglendi <- spatSample(skoglendi, size = 5000, method = "random")
pts_birki <- spatSample(natturu_birki, size = 5000, method = "random")

# --- 3) Combine both point sets ---
pts_combined <- rbind(pts_skoglendi, pts_birki)

# --- 4) Check result ---
nrow(pts_combined)  # should be 4000

# --- 5) Plot to verify visually ---
plot(skoglendi, border = "darkgreen", col = "lightgreen")
plot(natturu_birki, border = "brown", add = TRUE)
points(pts_combined, col = "red", pch = 16)

iceland_crs <- "+proj=lcc +lat_0=65 +lon_0=-19 +lat_1=64.25 +lat_2=65.75 +x_0=2700000 +y_0=300000 +ellps=GRS80 +units=m +no_defs"
presence_locations <- project(pts_combined, iceland_crs)

```

## Sampling psuedo-absences for SDM from habitat map

The idea here it to

```{r}

# --- Birch habitat suitability weights (0–1 scale) ---
# Natural types = positive weights
# Artificial / built-up / roads = 0

birch_weights <- data.frame(
  Value = 1:65,  # expanded to include anthropogenic types
  Vistgerd = c(
    "L1.1 Eyðimelavist",
    "L1.2 Grasmelavist",
    "L1.3 Mosamelavist",
    "L1.4 Víðimelavist",
    "L1.5 Sanda- og vikravist",
    "L3.3 Ljónslappaskriðuvist",
    "L3.1 Urðarskriðuvist",
    "L3.2 Grasvíðiskriðuvist",
    "L4.2 Auravist",
    "L4.1 Eyravist",
    "L5.1 Hélumosavist",
    "L5.2 Melagambravist",
    "L5.3 Hraungambravist",
    "L6.2 Fléttuhraunavist",
    "L6.1 Eyðihraunavist",
    "L6.4 Lynghraunavist",
    "L6.3 Mosahraunavist",
    "L2.1 Moldavist",
    "L7.7 Sjávarkletta- og eyjavist",
    "L7.4 Grashólavist",
    "L7.6 Gulstararfitjavist",
    "L7.3 Strandmelhólavist",
    "L7.5 Sjávarfitjungsvist",
    "L7.2 Malarstrandarvist",
    "L7.1 Sandstrandarvist",
    "L8.1 Dýjavist",
    "L8.14 Gulstararflóavist",
    "L8.4 Hrossanálarvist",
    "L8.11 Brokflóavist",
    "L8.10 Hengistararflóavist",
    "L8.12 Starungsflóavist",
    "L8.9 Starungsmýravist",
    "L8.2 Rekjuvist",
    "L8.6 Runnamýravist á láglendi",
    "L8.5 Runnamýravist á hálendi",
    "L8.8 Rústamýravist",
    "L8.3 Sandmýravist",
    "L8.13 Tjarnastararflóavist",
    "L9.7 Blómgresisvist",
    "L9.3 Bugðupuntsvist",
    "L9.5 Grasengjavist",
    "L9.2 Finnungsvist",
    "L9.6 Língresis- og vingulsvist",
    "L9.4 Snarrótarvist",
    "L9.1 Stinnastararvist",
    "L10.6 Fjalldrapamóavist",
    "L10.5 Fléttumóavist",
    "L10.4 Grasmóavist",
    "L10.7 Lyngmóavist á hálendi",
    "L10.8 Lyngmóavist á láglendi",
    "L10.1 Mosamóavist",
    "L10.2 Flagmóavist",
    # --- Anthropogenic categories ---
    "L11.1 Ræktarlönd",               # cultivated/agricultural land
    "L11.2 Túngrónir reitir",          # hayfields/pastures
    "L12.1 Vegsvæði",                 # roads
    "L12.2 Þéttbýli / byggð",         # built-up area
    "L12.3 Flugvellir / iðnaðarsvæði",# airports / industry
    "L12.4 Námur",                    # quarries
    "L12.5 Vökvafylltar lón",         # reservoirs
    "L12.6 Skógræktarsvæði",          # forest plantations (may have birch)
    
    "L14.1 Þéttbýli og annað manngert land",
    "L14.4 Alaskalúpína",
    "V1 Vötn",
    "L11 Birkiskógur",
    "L14.3 Skógrækt"
    	
  ),
  Birch_weight = c(
    0.05, 0.10, 0.30, 0.10, 0.05, 0.15, 0.10, 0.20, 0.05, 0.05,
    0.40, 0.45, 0.50, 0.55, 0.30, 0.65, 0.55, 0.35, 0.05, 0.10,
    0.05, 0.05, 0.05, 0.05, 0.05, 0.40, 0.45, 0.35, 0.50, 0.45,
    0.50, 0.50, 0.40, 0.55, 0.50, 0.45, 0.40, 0.45, 0.70, 0.65,
    0.70, 0.65, 0.70, 0.65, 0.60, 0.85, 0.75, 0.80, 0.40, 0.20,
    0.75, 0.60,
    # Anthropogenic weights
    0.00,  # L11.1 Ræktarlönd
    0.10,  # L11.2 Túngrónir reitir (some birch along edges)
    0.00,  # L12.1 Vegsvæði
    0.00,  # L12.2 Þéttbýli / byggð
    0.00,  # L12.3 Flugvellir / iðnaðarsvæði
    0.00,  # L12.4 Námur
    0.00,  # L12.5 Vökvafylltar lón (water)
    0.00,  # L12.6 Skógræktarsvæði (birch reforestation possible)
    0.00,
    0.00,
    0.00,
    0.00,
    0.00
  )
)

# 1. Load raster
habitats <- rast("data/natt_vg25r_3_utg_epsg_3057/NI_VG25r_3.utg/ni_vg25r_3utg.tif")

# 2. Read attribute table
rat <- as.data.frame(read.dbf("data/natt_vg25r_3_utg_epsg_3057/NI_VG25r_3.utg/ni_vg25r_3utg.tif.vat.dbf"))

# 3. Join with birch weights
rat_joined <- rat %>%
  dplyr::rename(Vistgerd = dbf.Vistgerð, Value = dbf.Value) %>%
  left_join(birch_weights, by = "Vistgerd")

# 4. Check join result
head(rat_joined[c("Value.x", "Vistgerd", "Birch_weight")])

# 5. Prepare lookup matrix for classify()
lookup_mat <- as.matrix(rat_joined[, c("Value.x", "Birch_weight")])

# 6. Map birch weights onto raster
habitat_birch_suitability <- classify(habitats, rcl = lookup_mat)

```

```{r include = F}

# Each 1 km cell is 1000 / 10 = 100 cells of 10 m
fact <- 100  # aggregation factor

habitat_birch_suitability_1km <- aggregate(
  habitat_birch_suitability,
  fact = fact,
  fun = mean,
  na.rm = T
)

# --- Generate pseudo-absences across Iceland ---

set.seed(456)

# 1. Raster to data.frame
r_df <- as.data.frame(habitat_birch_suitability_1km, xy = TRUE, na.rm = TRUE)
names(r_df) <- c("x", "y", "suitability")

# 2. Remove zero values (optional, or replace with small epsilon)
r_df <- r_df[r_df$suitability > 0, ]

# 3. Weighted sampling of rows
sample_idx <- sample(
  1:nrow(r_df),
  size = 5000,
  replace = TRUE,            # TRUE allows sampling more points than unique cells
  prob = r_df$suitability    # probability proportional to raster value
)

sampled_points_df <- r_df[sample_idx, c("x", "y")]

# 4. Convert to SpatVector
pseudo_absences <- vect(sampled_points_df, geom = c("x", "y"), crs = crs(habitat_birch_suitability_1km))

plot(habitat_birch_suitability_1km)
points(pseudo_absences, col = "red", pch = 16, cex = 0.6)

pseudo_absences <- project(pseudo_absences, iceland_crs)

```

```{r include = F}


pseudo_absences$presence <- 0  # mark pseudo-absences
presence_locations$presence <- 1

# Keep only presence column for combining
presence_locations <- presence_locations["presence"]
pseudo_absences   <- pseudo_absences["presence"]

# --- 6) Combine presence + pseudo-absence points ---
pa_data <- rbind(presence_locations, pseudo_absences)


```

### Regional-scale SDM

load in the soil


```{r include = F}
# Reading in bioclim data for a cropped and masked Iceland

data_folder <- "data/wc2.1_30s_bio/"
files <- list.files(
  data_folder,
  pattern = "^wc2\\.1_30s_bio_.*\\.tif$", 
  full.names = TRUE
)

bio_stack <- rast(files)

iceland_shape <- vect("data/Iceland_shapefile/is_1km.shp")
iceland_shape <- project(iceland_shape, crs(bio_stack))
bio_iceland <- crop(bio_stack, iceland_shape)
bio_regional_crop <- mask(bio_iceland, iceland_shape)

```

```{r include = F}
# Reading in DEM to crop and mask and calculate slope

dem_regional <-  rast('data/wc2.1_30s_elev/wc2.1_30s_elev.tif')
dem_regional_crop <- crop(dem_regional, iceland_shape)
dem_regional_crop <- mask(dem_regional_crop, iceland_shape)
slope_regional <- terrain(dem_regional_crop, v = "slope", unit = "degrees")

```

```{r include = F}
# Topographic position index of 5 x 5 moving window using mean

w <- matrix(1, nrow = 5, ncol = 5)
local_mean <- focal(dem_regional_crop, w = w, fun = mean, na.rm = TRUE)
tpi <- dem_regional_crop - local_mean
names(tpi) <- "tpi"

```


```{r include = F}
# compile raster stack for modelling

soil_stack <- rast("data/soils/iceland_soilgrid_0-5cm.tif")
soil_stack <- project(soil_stack, bio_stack)
soil_stack <- crop(soil_stack, iceland_shape)

```

```{r include = F}
# compile raster stack for modelling

env_stack <- c(bio_regional_crop, slope_regional, tpi, soil_stack)

```

```{r include = F}
# Assigning presence and absence

pa_climate <- terra::extract(env_stack, pa_data, xy = TRUE)
pa_climate$presence <- pa_data$presence
pa_climate <- na.omit(pa_climate)
pa_climate$presence <- factor(pa_climate$presence, levels = c(0,1))
pa_climate <- pa_climate %>% dplyr::select(-c(ID,x,y))


```

```{r include = F}
# fitting the model

set.seed(123)
rf_broad_model <- randomForest(
presence ~ .,
data = pa_climate,
ntree = 500,
mtry = floor(sqrt(ncol(pa_climate))),  # fewer variables per split
nodesize = 10,  
importance = TRUE
)

importance(rf_broad_model)
varImpPlot(rf_broad_model, type = 1)
```


### Cross validation

Spatial cross-validation (CV) is a method used to evaluate species distribution models while accounting for spatial autocorrelation in ecological data. Traditional random CV can overestimate model performance because nearby locations often have similar environmental conditions, causing the training and test data to be spatially dependent. In spatial CV, the study area is divided into spatial blocks, and each block is systematically held out as a test set while the model is trained on the remaining blocks. This approach ensures that training and testing data are spatially separated, providing a more realistic assessment of how well the model can predict species presence or absence in unsampled areas. By repeating this process across multiple blocks or folds, researchers can obtain robust performance metrics such as accuracy, sensitivity, specificity, and AUC, giving a more reliable estimate of model generalization in space.

```{r}

# 1) Convert presence/absence points to sf and project to raster CRS
pa_sf <- st_as_sf(pa_data)  
pa_sf <- st_transform(pa_sf, st_crs(env_stack))

# 2) Spatial blocking
set.seed(123)
sb <- spatialBlock(
  speciesData = pa_sf,
  species = "presence",
  rasterLayer = env_stack[[1]],  # defines extent
  theRange = 50000,              # block size in meters
  k = 5,                         # number of folds
  selection = "random",
  iteration = 100,
  showBlocks = TRUE
)
pa_sf$foldID <- sb$foldID

# 3) Split into training/testing for fold 1
train <- pa_sf[pa_sf$foldID != 1, ]
test  <- pa_sf[pa_sf$foldID == 1, ]

# 4) Convert sf to SpatVector for terra::extract
train_vect <- vect(train)
test_vect  <- vect(test)

# 5) Extract environmental variables and bind presence column
train_env <- terra::extract(env_stack, train_vect, bind = TRUE)
test_env  <- terra::extract(env_stack, test_vect, bind = TRUE)

# 6) Convert to data.frame for dplyr manipulation
train_env <- as.data.frame(train_env)
test_env  <- as.data.frame(test_env)

# 7) Remove ID column if it exists, then remove rows with NA
cols_to_remove <- intersect(c("ID"), colnames(train_env))
if(length(cols_to_remove) > 0) train_env <- train_env %>% select(-all_of(cols_to_remove))
train_env <- na.omit(train_env)

cols_to_remove <- intersect(c("ID"), colnames(test_env))
if(length(cols_to_remove) > 0) test_env <- test_env %>% select(-all_of(cols_to_remove))
test_env <- na.omit(test_env)

# 8) Convert presence to factor
train_env$presence <- factor(train_env$presence, levels = c(0,1))
test_env$presence  <- factor(test_env$presence, levels = c(0,1))

# 9) Quick NA check
colSums(is.na(train_env))  # should be all 0
colSums(is.na(test_env))   # should be all 0

# 10) Fit Random Forest
rf_model <- randomForest(
  presence ~ .,
  data = pa_climate,
  ntree = 500,
  mtry = floor(sqrt(ncol(pa_climate))),  # fewer variables per split
  nodesize = 10,  
  importance = TRUE)

# 11) Predict on test set
pred <- predict(rf_model, newdata = test_env, type = "response")

# 12) Evaluate performance
confusionMatrix(pred, test_env$presence)

```

### Prediction

```{r}
# predicting over Iceland

broad_scale_rf_prob <- terra::predict(
  env_stack,             # climate predictors raster stack
  rf_broad_model,                # your trained RF model
  type = "prob",
  index = 2, # probability of presence class
  filename = "data/broad_scale_rf_prob_birch.tif",  # write output directly to file
  overwrite = TRUE)

```

### smooths edges

Since we are combining broad-scale and local-scale models, we need to smooth the edges of the broad-scale model so that when we clip to the AOI there are no jagged edges. 

```{r}

# Broadscale model (already loaded)
broad_prob <- broad_scale_rf_prob

# Reproject to a metric CRS (EPSG:8088 – Iceland Lambert 2016)
broad_proj <- terra::project(broad_prob, "EPSG:8088")

# Fill small NA gaps using a local mean
broad_filled <- terra::focal(
  broad_proj,
  w = 3,
  fun = function(x, ...) {
    if (is.na(x[5])) mean(x, na.rm = TRUE) else x[5]
  },
  na.policy = "omit",
  fillvalue = NA
)

# Define a 5 km smoothing radius
radius_m <- 5000
w <- terra::focalMat(broad_filled, d = radius_m, type = "circle")

# Apply smoothing (mean filter)
broad_smooth <- terra::focal(broad_filled, w = w, fun = mean, na.rm = TRUE)

# Reproject smoothed raster back to WGS84 (or original CRS of broad_prob)
broad_smooth_aligned <- terra::project(broad_smooth, crs(broad_prob), method = "bilinear")

# Resample smoothed raster to exactly match broad_prob (extent & resolution)
broad_smooth_aligned <- terra::resample(broad_smooth_aligned, broad_prob, method = "bilinear")

# Cover edges: keep broad_prob, fill NAs with smoothed raster
broad_edges_added <- terra::cover(broad_prob, broad_smooth_aligned)


```


```{r}
dem <- rast("data/IslandsDEMv1.0_10x10m_isn2016_zmasl.tif")

broad_scale_rf_prob_10m <- terra::project(broad_edges_added, dem)

```


```{r}

plot(broad_scale_rf_prob_10m)

```

```{r}
# ---- Predict probability of presence at each point used in training ----
pa_climate$pred_prob <- predict(rf_broad_model, type = "prob")[, 2]  # column 2 = presence class prob

# ---- Compute residuals ----
# presence is coded as 0 (pseudo-absence) or 1 (presence)
# residual = observed - predicted
pa_climate$residual <- as.numeric(as.character(pa_climate$presence)) - pa_climate$pred_prob

```

```{r}

# Convert residuals to SpatVector with coordinates
resid_vect <- pa_data  # same geometry as presence/pseudoabsence points
resid_vect$residual <- pa_climate$residual


```

## GEE 

This code chunk outputs the points needed for GEE.


```{r eval = F}
# export points for future work with GEE - Multiple mechanisms of early plant community assembly with stochasticity driving the process

sampled_forest_cells <- project(pa_data[, FALSE], "EPSG:4326")
writeVector(sampled_forest_cells, "data/GEE/SDM_pa_points_birch.shp", overwrite = TRUE)

```

### Fine-scale SDM

## Construct fine-scale predictors

## Elevation and slope

```{r}

# Compute slope and aspect
slope <- terrain(dem, v = "slope", unit = "radians")

```

```{r}

dem_100m <- aggregate(dem, fact = 10, fun = mean)  # 100 m resolution
writeRaster(dem_100m, filename = "data/saga_output/DEM_100m.tif", overwrite = TRUE)

```

## Mean growing season and peak summer radiation 

```{r}

# set environment (adjust path if needed)
env <- rsaga.env(path = "C:/Users/jay.philip/Desktop/utilities/saga-9.10.0_x64/saga-9.10.0_x64")   # auto-detects SAGA; or rsaga.env(path = "C:/path/to/saga")
outdir <- "data/saga_output/"

# -----------------------
# 3. Define representative mid-month days
# -----------------------
# List of months and representative day of month
mid_month_days <- data.frame(
  month = 5:9,         # May–September
  day   = c(15, 15, 15, 15, 15)
)

# Store paths to Total radiation rasters
total_rasters <- c()
jun_jul_rasters <- c()

# -----------------------
# 4. Loop over each mid-month day
# -----------------------
for(i in 1:nrow(mid_month_days)) {
  m <- mid_month_days$month[i]
  d <- mid_month_days$day[i]
  
  out_total <- file.path(outdir, paste0("Total_", sprintf("%02d", m), ".tif"))
  out_direct <- file.path(outdir, paste0("Direct_", sprintf("%02d", m), ".tif"))
  out_diffuse <- file.path(outdir, paste0("Diffuse_", sprintf("%02d", m), ".tif"))
  out_duration <- file.path(outdir, paste0("Duration_", sprintf("%02d", m), ".tif"))
  
  cat("Computing solar radiation for month", m, "day", d, "\n")
  
  rsaga.pisr2(
    in.dem         = "data/saga_output/DEM_100m.tif",
    out.total.grid = out_total,
    out.direct.grid = out_direct,
    out.diffuse.grid = out_diffuse,
    out.duration    = out_duration,
    latitude       = 65,               # Iceland
    unit           = "kWh/m2",
    time.range     = c(0, 24),
    time.step      = 1,                # hourly integration
    start.date     = list(day = d, month = m, year = 2024),
    day.step       = 1,
    env            = env
  )
  
  # Collect raster paths for later averaging
  total_rasters <- c(total_rasters, out_total)
  if(m %in% 6:7) {  # June–July for peak summer
    jun_jul_rasters <- c(jun_jul_rasters, out_total)
  }
}

# -----------------------
# 5. Aggregate Total radiation rasters
# -----------------------

# Stack all Total rasters into a single SpatRaster for growing season
rast_stack <- do.call(c, lapply(total_rasters, rast))
growing_season_mean <- app(rast_stack, mean, na.rm = TRUE)
writeRaster(growing_season_mean, file.path(outdir, "Mean_GrowingSeason.tif"), overwrite = TRUE)

# Stack June–July rasters for peak summer
jun_jul_stack <- do.call(c, lapply(jun_jul_rasters, rast))
peak_summer_mean <- app(jun_jul_stack, mean, na.rm = TRUE)
writeRaster(peak_summer_mean, file.path(outdir, "Mean_JuneJuly.tif"), overwrite = TRUE)

cat("Finished computing growing season and peak summer radiation rasters.\n")


```

```{r}

# -----------------------
# 1. Read the averaged rasters
# -----------------------
growing_season <- rast(file.path(outdir, "Mean_GrowingSeason.tif"))
peak_summer   <- rast(file.path(outdir, "Mean_JuneJuly.tif"))

# -----------------------
# 2. Plot side by side
# -----------------------
par(mfrow = c(1, 2))  # two plots in one row

plot(growing_season, main = "Mean Growing Season Radiation\n(May–Sep, kWh/m²/day)")
plot(peak_summer, main = "Peak Summer Radiation\n(June–July, kWh/m²/day)")

par(mfrow = c(1,1))  # reset plotting layout


```

```{r}

# Upscale mean rasters to 10m
growing_season_10m <- resample(growing_season, dem, method = "bilinear")
peak_summer_10m   <- resample(peak_summer, dem, method = "bilinear")

# Write out upscaled rasters
writeRaster(growing_season_10m, file.path(outdir, "Mean_GrowingSeason_10m.tif"), overwrite = TRUE)
writeRaster(peak_summer_10m,   file.path(outdir, "Mean_JuneJuly_10m.tif"), overwrite = TRUE)

```

## Topographic wetness index

```{r eval=F}

# -----------------------
# SDM Topographic Predictor Preparation
# Computes: slope, aspect, upslope area (SCA), TWI
# -----------------------

library(RSAGA)
library(terra)

# -----------------------
# 1. Paths and environment
# -----------------------
dem_path <- "data/IslandsDEMv1.0_10x10m_isn2016_zmasl.tif"
dir.create(outdir, showWarnings = FALSE)

# -----------------------
# 2. Slope & Aspect
# -----------------------
slope_out  <- file.path(outdir, "Slope.sgrd")
aspect_out <- file.path(outdir, "Aspect.sgrd")

rsaga.geoprocessor(
  lib    = "ta_morphometry",
  module = "Slope, Aspect, Curvature",
  param  = list(
    ELEVATION = dem_path,
    SLOPE     = slope_out,
    ASPECT    = aspect_out
  ),
  env = env,
  show.output.on.console = TRUE
)

# -----------------------
# 3. Upslope Area (Specific Catchment Area)
# -----------------------
sca_out <- file.path(outdir, "SCA.sgrd")

rsaga.geoprocessor(
  lib    = "ta_hydrology",
  module = "Flow Width and Specific Catchment Area",
  param  = list(
    DEM        = dem_path,
    WIDTH      = NULL,        # optional
    TCA        = NULL,        # optional
    SCA        = sca_out,     # output specific catchment area
    COORD_UNIT = 0,
    METHOD     = 0
  ),
  env = env,
  show.output.on.console = TRUE
)

# -----------------------
# 4. TWI Calculation
# -----------------------
twi_out <- file.path(outdir, "TWI.sgrd")

rsaga.geoprocessor(
  lib    = "ta_hydrology",
  module = "Topographic Wetness Index",
  param  = list(
    SLOPE  = slope_out,
    AREA   = sca_out,   # use SCA for TWI
    TRANS  = 0,
    TWI    = twi_out,
    CONV   = 1,
    METHOD = 0
  ),
  env = env,
  show.output.on.console = TRUE
)

```

```{r}

twi <- rast("data/saga_output/TWI.tif")

```

## Modelling residuals experimental

```{r}
library(terra)
library(sf)
library(dplyr)
library(xgboost)

# ===========================
# 1. Raster predictors
# ===========================
env_stack <- c(dem, slope, peak_summer_10m, twi)
names(env_stack) <- c("DEM", "Slope", "peak_summer", "twi")

# Extract raster values at residual points
resid_env <- terra::extract(env_stack, resid_vect, xy = TRUE)

# Add residuals from resid_vect
resid_env$residual <- resid_vect$residual

# Remove incomplete rows
resid_env <- resid_env %>% dplyr::filter(dplyr::if_all(dplyr::everything(), ~ !is.na(.)))

# ===========================
# 2. Embeddings
# ===========================
embeddings <- read.csv("data/GEE/birch_embeddings_2023_10m_avg.csv") %>%
  dplyr::rename(x = longitude, y = latitude) %>%
  dplyr::select(-dplyr::any_of(c("FID", "system.index","year", ".geo")))

# Convert to sf
embeddings_sf <- sf::st_as_sf(embeddings, coords = c("x", "y"), crs = 4326) # assume embeddings in WGS84

# Convert residuals to sf
resid_sf <- sf::st_as_sf(resid_vect)

# ===========================
# 3. CRS alignment
# ===========================
# Transform embeddings to match residuals CRS
embeddings_sf <- sf::st_transform(embeddings_sf, sf::st_crs(resid_sf))

# ===========================
# 4. Spatial join (nearest neighbor)
# ===========================
resid_with_embed_sf <- sf::st_join(resid_sf, embeddings_sf, join = sf::st_nearest_feature)

# Drop geometry
resid_with_embed_df <- sf::st_drop_geometry(resid_with_embed_sf)

# ===========================
# 5. Combine raster predictors + embeddings
# ===========================
resid_env_sf <- sf::st_as_sf(resid_env, coords = c("x", "y"), crs = sf::st_crs(resid_sf))

# Join raster predictors to embeddings
full_sf <- sf::st_join(resid_env_sf, embeddings_sf, join = sf::st_nearest_feature)

# Drop geometry to get final data frame
full_df <- sf::st_drop_geometry(full_sf)

# Verify residual column exists
if(!"residual" %in% colnames(full_df)) stop("Column 'residual' not found in full_df")

# ===========================
# 6. Prepare for XGBoost
# ===========================
resid_mat <- full_df %>%
  dplyr::select(-dplyr::any_of(c("residual","ID"))) %>%
  as.matrix()

resid_target <- full_df$residual

dtrain <- xgb.DMatrix(data = resid_mat, label = resid_target)

# ===========================
# 7. Train model
# ===========================
params <- list(
  objective = "reg:squarederror",
  eta = 0.1,
  max_depth = 4,
  subsample = 0.7,
  colsample_bytree = 0.7
)

set.seed(123)
xgb_fine_model_res <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 500,
  verbose = 1
)

# ===========================
# 8. Feature importance
# ===========================
importance_matrix <- xgb.importance(model = xgb_fine_model_res)
xgb.plot.importance(importance_matrix)


```

## Export AOI for GEE

```{r}

# Load regions
regions <- vect("data/boundaries/is_50v_mork_epsg_8088.gpkg/is_50v_mork_epsg_8088.gpkg", 
                layer = "mork_logregluumdaemi")

# Reproject to DEM CRS
regions_proj <- project(regions, crs(dem))

# Subset the region you want (2) reyjkanes
rykb_subset <- regions_proj[regions_proj$objectid == 3,]

# Convert to sf polygon (for GEE)
rykb_sf <- st_as_sf(rykb_subset)

# Split MULTIPOLYGON into individual POLYGON rows------
rykb_single_polys <- st_cast(rykb_sf, "POLYGON")
rykb_single_polys <- rykb_single_polys %>%
  mutate(area_m2 = st_area(geometry))

rykb_ordered <- rykb_single_polys %>%
  arrange(desc(area_m2))

second_largest_poly <- rykb_ordered[2, ]  # 2nd row after sorting
second_largest_poly <- second_largest_poly %>%
  select(-area_m2)

### combine it with reyjkanes

rykb_subset <- regions_proj[regions_proj$objectid == 2,]
rykb_sf <- st_as_sf(rykb_subset)

#merge 
merged_sf <- rbind(rykb_sf, second_largest_poly)
rykb_subset <- st_union(merged_sf)


# Save as GeoJSON for GEE upload
# st_write(rykb_subset, "data/GEE/complete_reykjanes_penisula.shp", delete_dsn = TRUE, overwrite = T)

```

## Read in Embeddings

```{r}

library(terra)

# Set directory
raster_dir <- "data/GEE/reyk_penisula_tiles"

# List all TIFF files
tif_files <- list.files(raster_dir, pattern = "\\.tif$", full.names = TRUE)

# Read all rasters into a list
rasters_list <- lapply(tif_files, rast)

# Merge tiles together (works even if extents don't match)
merged_raster <- do.call(merge, rasters_list)

# Check merged raster
print(merged_raster)
plot(merged_raster[[30]])  # plot first layer as example

```

## Create fine-scale topographic-embeddings rasterstack

```{r}

env_stack <- crop(env_stack, rykb_subset)
# env_stack <- mask(env_stack, rykb_subset)

# 1️⃣ Align CRS of merged_raster to env_stack
merged_raster_proj <- project(merged_raster, crs(env_stack))

# 2️⃣ Resample merged_raster to match env_stack resolution and extent
merged_raster_resampled <- resample(merged_raster_proj, env_stack, method = "bilinear") 
# "bilinear" is good for continuous data; use "near" for categorical

# 3️⃣ Crop to exact env_stack extent (optional, just to be safe)
merged_raster_crop <- crop(merged_raster_resampled, ext(env_stack))

# 4️⃣ Stack with existing env_stack
full_stack <- c(env_stack, merged_raster_crop)

# 5️⃣ Verify
names(full_stack)
plot(full_stack[[1]])  # Check first layer


```

## Predict over AOI

```{r}
library(terra)

# Extract predictor values from raster for Reykjanes
pred_mat <- terra::values(full_stack, mat = TRUE)  # matrix of raster values

# Remove rows with NA (XGBoost cannot handle NAs)
na_rows <- apply(pred_mat, 1, function(x) any(is.na(x)))
pred_mat_clean <- pred_mat[!na_rows, , drop = FALSE]

# Keep track of which cells are valid
valid_cells <- which(!na_rows)

# Predict residuals
resid_pred <- predict(
  xgb_fine_model_res,
  newdata = pred_mat_clean
)

# Create an empty raster to store predictions
resid_raster <- full_stack[[1]]  # template
terra::values(resid_raster) <- NA
terra::values(resid_raster)[valid_cells] <- resid_pred

# Plot
plot(resid_raster, main = "Fine-scale Residual Prediction (Gradient Boosting)")
terra::writeRaster(resid_raster, "data/residual_plot_embedded.tif", overwrite = TRUE)

```

## Resample broad-scale model

```{r}

# Ensure CRS match
broad_scale_rf_prob_10m <- project(broad_scale_rf_prob_10m, crs(rykb_subset))

# Crop to the Reykjanes bounding box
broad_crop <- crop(broad_scale_rf_prob_10m, rykb_subset)
# broad_crop <- mask(broad_crop, rykb_subset)

plot(broad_crop, main = "Broad-scale probability (Reykjanes subset)")

```

```{r}

writeRaster(broad_crop, "data/broad_scale_model_cropped.tif", overwrite=TRUE)

```


## Combine residuals to broad-scale probabilities and construct final predictions

```{r}

# Align grid, extent, and resolution
fine_resid_pred <- project(resid_raster, crs(broad_crop))
fine_resid_pred <- resample(resid_raster, broad_crop, method = "bilinear")

# Combine
multiscale_reykjanes <- broad_crop + fine_resid_pred

# Clamp to valid probability range
multiscale_reykjanes <- clamp(multiscale_reykjanes, 0, 1)

# Plot
plot(multiscale_reykjanes, main = "Multiscale Betula SDM (Reykjanes, 10 m)")

# Save output
writeRaster(multiscale_reykjanes,
            "data/multiscale_reykjanes_10m_embeddings_predictions.tif",
            overwrite = TRUE)

```

