---
title: "Afforestation planning in Iceland"
author: "J.P.Borchard"
date: "2025-10-16"
output: html_document
---

### Introduction




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include = F}

library(raster)
library(terra)
library(sf)
library(dplyr)
library(leaflet)
library(whitebox)
library(RSAGA)
library(randomForest)
library(foreign)
library(blockCV)
library(caret)
library(spdep)
library(terra)
library(sf)
library(dplyr)
library(xgboost)

```

```{r include = F}
whitebox::wbt_init()


```

```{r include = F}
# lo
iceland_shape <- vect("data/Iceland_shapefile/is_1km.shp")
iceland_crs <- "+proj=lcc +lat_0=65 +lon_0=-19 +lat_1=64.25 +lat_2=65.75 +x_0=2700000 +y_0=300000 +ellps=GRS80 +units=m +no_defs"

```

### Sampling Birch Presences and Pseudo-Absences for Broad-Scale SDM – Iceland

 Overview:
 This workflow constructs the presence–absence dataset used to train the
 broad-scale birch Species Distribution Model (SDM) for Iceland. It combines
 (1) field-derived birch presence locations from national vegetation maps and
 (2) ecologically informed pseudo-absences sampled using a habitat-weighted
 probability surface derived from the NI vegetation classification.

 Workflow:

 1. Birch Presences:
    - The national “Náttúrulegt birkilendi” dataset is projected to a common CRS.
    - A random spatial sample of birch presence points is drawn from within
      mapped birch habitat polygons (e.g., 5,000 presences).

 2. Habitat Suitability Weights (0–1):
    - A lookup table is created that assigns each NI vegetation type
      (“Vistgerðir”) a birch suitability value between 0 and 1.
    - Natural habitats receive positive weights; anthropogenic classes
      (urban, roads, quarries, reservoirs, etc.) receive weight = 0.
    - This table is joined to the vegetation raster attribute table.

 3. Build a Habitat-Weighted Sampling Surface:
    - The NI vegetation raster is classified into a “birch suitability raster”
      using the weighted lookup table.
    - This 10 m suitability raster is aggregated to 1 km resolution to create
      a smooth landscape-level probability surface for sampling absences.

 4. Pseudo-Absence Sampling:
    - All non-NA 1 km cells are converted to a data frame.
    - Pseudo-absences are randomly drawn with probability proportional to the
      birch suitability value of each cell. This ensures:
         • Fewer pseudo-absences in highly unsuitable areas (urban, water).
         • More pseudo-absences in plausible birch environments, avoiding
           unrealistic contrasts.
    - A SpatVector of pseudo-absence points is created and projected to the
      Iceland CRS.

 5. Combine Presences and Pseudo-Absences:
    - Presence points are labelled presence = 1.
    - Pseudo-absence points are labelled presence = 0.
    - Both sets are merged into a single training dataset for SDM modelling.

 Features:
 - Ecologically informed pseudo-absence generation avoids unrealistic “true”
   absences and follows best practices for SDM pseudo-absence sampling.
 - Sampling is weighted, not random: pseudo-absences reflect habitat structure.
 - Spatial consistency is maintained by projecting all layers to a common CRS.

 Output:
 A SpatVector `pa_data` containing:
    • x, y coordinates of points
    • presence = 1 (birch) or 0 (pseudo-absence)

 This dataset is then used to train the broad-scale SDM before residual
 modelling and fine-scale correction.

```{r include = F}
set.seed(123)

natturu_birki <- project(vect("data/forests/Natturulegt_birkilendi_Island_15052025.shp"), iceland_crs)
presence_locations  <- spatSample(natturu_birki, size = 5000, method = "random")
```

```{r}

# --- Birch habitat suitability weights (0–1 scale) ---
# Natural types = positive weights
# Artificial / built-up / roads = 0

birch_weights <- data.frame(
  Value = 1:65,  # expanded to include anthropogenic types
  Vistgerd = c(
    "L1.1 Eyðimelavist",
    "L1.2 Grasmelavist",
    "L1.3 Mosamelavist",
    "L1.4 Víðimelavist",
    "L1.5 Sanda- og vikravist",
    "L3.3 Ljónslappaskriðuvist",
    "L3.1 Urðarskriðuvist",
    "L3.2 Grasvíðiskriðuvist",
    "L4.2 Auravist",
    "L4.1 Eyravist",
    "L5.1 Hélumosavist",
    "L5.2 Melagambravist",
    "L5.3 Hraungambravist",
    "L6.2 Fléttuhraunavist",
    "L6.1 Eyðihraunavist",
    "L6.4 Lynghraunavist",
    "L6.3 Mosahraunavist",
    "L2.1 Moldavist",
    "L7.7 Sjávarkletta- og eyjavist",
    "L7.4 Grashólavist",
    "L7.6 Gulstararfitjavist",
    "L7.3 Strandmelhólavist",
    "L7.5 Sjávarfitjungsvist",
    "L7.2 Malarstrandarvist",
    "L7.1 Sandstrandarvist",
    "L8.1 Dýjavist",
    "L8.14 Gulstararflóavist",
    "L8.4 Hrossanálarvist",
    "L8.11 Brokflóavist",
    "L8.10 Hengistararflóavist",
    "L8.12 Starungsflóavist",
    "L8.9 Starungsmýravist",
    "L8.2 Rekjuvist",
    "L8.6 Runnamýravist á láglendi",
    "L8.5 Runnamýravist á hálendi",
    "L8.8 Rústamýravist",
    "L8.3 Sandmýravist",
    "L8.13 Tjarnastararflóavist",
    "L9.7 Blómgresisvist",
    "L9.3 Bugðupuntsvist",
    "L9.5 Grasengjavist",
    "L9.2 Finnungsvist",
    "L9.6 Língresis- og vingulsvist",
    "L9.4 Snarrótarvist",
    "L9.1 Stinnastararvist",
    "L10.6 Fjalldrapamóavist",
    "L10.5 Fléttumóavist",
    "L10.4 Grasmóavist",
    "L10.7 Lyngmóavist á hálendi",
    "L10.8 Lyngmóavist á láglendi",
    "L10.1 Mosamóavist",
    "L10.2 Flagmóavist",
    # --- Anthropogenic categories ---
    "L11.1 Ræktarlönd",               # cultivated/agricultural land
    "L11.2 Túngrónir reitir",          # hayfields/pastures
    "L12.1 Vegsvæði",                 # roads
    "L12.2 Þéttbýli / byggð",         # built-up area
    "L12.3 Flugvellir / iðnaðarsvæði",# airports / industry
    "L12.4 Námur",                    # quarries
    "L12.5 Vökvafylltar lón",         # reservoirs
    "L12.6 Skógræktarsvæði",          # forest plantations (may have birch)
    
    "L14.1 Þéttbýli og annað manngert land",
    "L14.4 Alaskalúpína",
    "V1 Vötn",
    "L11 Birkiskógur",
    "L14.3 Skógrækt"
    	
  ),
  Birch_weight = c(
    0.05, 0.10, 0.30, 0.10, 0.05, 0.15, 0.10, 0.20, 0.05, 0.05,
    0.40, 0.45, 0.50, 0.55, 0.30, 0.65, 0.55, 0.35, 0.05, 0.10,
    0.05, 0.05, 0.05, 0.05, 0.05, 0.40, 0.45, 0.35, 0.50, 0.45,
    0.50, 0.50, 0.40, 0.55, 0.50, 0.45, 0.40, 0.45, 0.70, 0.65,
    0.70, 0.65, 0.70, 0.65, 0.60, 0.85, 0.75, 0.80, 0.40, 0.20,
    0.75, 0.60,
    # Anthropogenic weights
    0.00,  # L11.1 Ræktarlönd
    0.10,  # L11.2 Túngrónir reitir (some birch along edges)
    0.00,  # L12.1 Vegsvæði
    0.00,  # L12.2 Þéttbýli / byggð
    0.00,  # L12.3 Flugvellir / iðnaðarsvæði
    0.00,  # L12.4 Námur
    0.00,  # L12.5 Vökvafylltar lón (water)
    0.00,  # L12.6 Skógræktarsvæði (birch reforestation possible)
    0.00,
    0.00,
    0.00,
    0.00,
    0.00
  )
)

# 1. Load raster
habitats <- rast("data/natt_vg25r_3_utg_epsg_3057/NI_VG25r_3.utg/ni_vg25r_3utg.tif")

# 2. Read attribute table
rat <- as.data.frame(read.dbf("data/natt_vg25r_3_utg_epsg_3057/NI_VG25r_3.utg/ni_vg25r_3utg.tif.vat.dbf"))

# 3. Join with birch weights
rat_joined <- rat %>%
  dplyr::rename(Vistgerd = dbf.Vistgerð, Value = dbf.Value) %>%
  left_join(birch_weights, by = "Vistgerd")

# 4. Check join result
head(rat_joined[c("Value.x", "Vistgerd", "Birch_weight")])

# 5. Prepare lookup matrix for classify()
lookup_mat <- as.matrix(rat_joined[, c("Value.x", "Birch_weight")])

# 6. Map birch weights onto raster
habitat_birch_suitability <- classify(habitats, rcl = lookup_mat)

```

```{r include = F}

# Each 1 km cell is 1000 / 10 = 100 cells of 10 m
fact <- 100  # aggregation factor

habitat_birch_suitability_1km <- aggregate(
  habitat_birch_suitability,
  fact = fact,
  fun = mean,
  na.rm = T
)

# --- Generate pseudo-absences across Iceland ---

set.seed(456)

# 1. Raster to data.frame
r_df <- as.data.frame(habitat_birch_suitability_1km, xy = TRUE, na.rm = TRUE)
names(r_df) <- c("x", "y", "suitability")

# 2. Remove zero values (optional, or replace with small epsilon)
r_df <- r_df[r_df$suitability > 0, ]

# 3. Weighted sampling of rows
sample_idx <- sample(
  1:nrow(r_df),
  size = 5000,
  replace = TRUE,            # TRUE allows sampling more points than unique cells
  prob = r_df$suitability    # probability proportional to raster value
)

sampled_points_df <- r_df[sample_idx, c("x", "y")]

# 4. Convert to SpatVector
pseudo_absences <- vect(sampled_points_df, geom = c("x", "y"), crs = crs(habitat_birch_suitability_1km))

plot(habitat_birch_suitability_1km)
points(pseudo_absences, col = "red", pch = 16, cex = 0.6)

pseudo_absences <- project(pseudo_absences, iceland_crs)

```

```{r include = F}


pseudo_absences$presence <- 0  # mark pseudo-absences
presence_locations$presence <- 1

# Keep only presence column for combining
presence_locations <- presence_locations["presence"]
pseudo_absences   <- pseudo_absences["presence"]

# --- 6) Combine presence + pseudo-absence points ---
pa_data <- rbind(presence_locations, pseudo_absences)
saveRDS(pa_data, "models/occurences.RDS")

```

### Broad-Scale Environmental Predictor Assembly – Iceland SDM

 Overview:
 This workflow prepares the environmental predictor stack used to train the
 broad-scale Iceland-wide birch SDM. It compiles climate, topographic, and soil
 layers, all resampled, cropped, and masked to a consistent Iceland extent
 and coordinate reference system.

 Workflow:

 1. Bioclimatic Predictors (WorldClim 2.1, 30 arc-sec):
    - All BIO climatic variables (BIO1–BIO19) are loaded.
    - Layers are cropped to a national Iceland boundary and masked to remove
      cells outside the country.
    - Provides temperature, precipitation, and seasonality predictors that
      capture coarse climatic gradients across Iceland.

 2. Elevation and Slope:
    - A 30 arc-sec DEM is loaded and aligned to the same extent as the BIO data.
    - Slope (in degrees) is derived using a terrain operator.
    - Elevation and slope together describe broad topographic variation, which
      strongly influences birch distribution (exposure, drainage, cold-air flow).

 3. Topographic Position Index (TPI):
    - A 5×5 moving window (mean filter) is used to compute local elevation.
    - TPI = elevation − local_mean, which captures whether a cell is:
         • ridge / convex (positive values),
         • flat / neutral (near zero),
         • valley / concave (negative values).
    - TPI acts as a coarse proxy for landform context and soil/moisture regimes.

 4. Soil Properties:
    - The upper soil layer (0–5 cm) from SoilGrids is projected to match the
      climatic layers and cropped to Iceland.
    - Soil variables (e.g., organic matter, pH, texture) act as slowly varying
      edaphic predictors for broad-scale birch distribution.

 5. Compile Final Predictor Stack:
    - All climate, topographic, and soil layers are concatenated into a single
      raster stack (`env_stack`) for SDM model fitting.
    - All predictors share:
         • identical CRS,
         • identical resolution,
         • identical extent,
         • full masking to Iceland boundaries.

 Features:
 - Integrates climate, terrain, and soils into a comprehensive broad-scale SDM
   predictor set.
 - Ensures spatial consistency across all layers via projection, cropping,
   masking, and alignment to the same template.
 - Designed to be coupled with presence/pseudo-absence data for XGBoost or
   other machine learning SDMs.

 Output:
 `env_stack` — a harmonised multi-layer raster stack containing all broad-scale
 predictors used for initial SDM model fitting before residual downscaling.


```{r include = F}
# Reading in bioclim data for a cropped and masked Iceland

data_folder <- "data/wc2.1_30s_bio/"
files <- list.files(
  data_folder,
  pattern = "^wc2\\.1_30s_bio_.*\\.tif$", 
  full.names = TRUE
)

bio_stack <- rast(files)

iceland_shape <- vect("data/Iceland_shapefile/is_1km.shp")
iceland_shape <- project(iceland_shape, crs(bio_stack))
bio_iceland <- crop(bio_stack, iceland_shape)
bio_regional_crop <- mask(bio_iceland, iceland_shape)

```

```{r include = F}
# Reading in DEM to crop and mask and calculate slope

dem_regional <-  rast('data/wc2.1_30s_elev/wc2.1_30s_elev.tif')
dem_regional_crop <- crop(dem_regional, iceland_shape)
dem_regional_crop <- mask(dem_regional_crop, iceland_shape)
slope_regional <- terrain(dem_regional_crop, v = "slope", unit = "degrees")

```

```{r include = F}
# Topographic position index of 5 x 5 moving window using mean

w <- matrix(1, nrow = 5, ncol = 5)
local_mean <- focal(dem_regional_crop, w = w, fun = mean, na.rm = TRUE)
tpi <- dem_regional_crop - local_mean
names(tpi) <- "tpi"

```


```{r include = F}
# compile raster stack for modelling

soil_stack <- rast("data/soils/iceland_soilgrid_0-5cm.tif")
soil_stack <- project(soil_stack, bio_stack)
soil_stack <- crop(soil_stack, iceland_shape)

```

```{r include = F}
# compile raster stack for modelling

env_stack <- c(bio_regional_crop, slope_regional, tpi, soil_stack)

```

### Broad-Scale Random Forest SDM Modelling – Iceland Birch Distribution

 Overview:
 This workflow trains the broad-scale Random Forest species distribution model
 (SDM) that estimates national-scale suitability for birch. Presence
 and pseudo-absence points are joined with the environmental predictor stack,
 cleaned, and used to fit a Random Forest classifier. The model captures coarse
 climatic and topographic controls on birch distribution across Iceland before
 fine-scale corrections are added via residual downscaling.

 Workflow:

 1. Extract Environmental Predictors at PA Locations:
    - Climate, terrain, and soil layers from env_stack are extracted for each
      presence and pseudo-absence location.
    - Rows containing missing values are removed to ensure a clean modelling
      dataset.
    - The response variable (presence) is encoded as a factor (0/1).
    - This produces the point-level environmental table used for model fitting.

 2. Fit the Broad-Scale Random Forest Model:
    - A Random Forest classifier is trained using:
         presence ~ all climate, topographic, and soil predictors
    - ntree = 500 ensures model stability, while mtry = sqrt(p) restricts the
      number of predictors randomly sampled at each split.
    - Variable importance metrics and importance plots identify the primary
      climatic drivers of birch distribution.

 3. Spatial Block Cross-Validation:
    - A 5-fold spatial block CV is constructed to avoid spatial autocorrelation
      inflating accuracy estimates.
    - Training and testing subsets are generated based on spatial folds rather
      than random point splits.
    - Model predictions for the held-out fold are compared against observed
      presence/absence to produce CV statistics and a confusion matrix.

 4. Iceland-Wide Prediction:
    - The trained Random Forest is projected across the full environmental
      predictor stack to generate a continuous probability-of-presence raster
      for all of Iceland.
    - The output raster is written directly to disk for downstream processing.

 5. Edge Smoothing and Reprojection:
    - To prevent sharp artefacts when combining broad- and local-scale
      predictions, the broad-scale raster is smoothed along boundaries using a
      5 km focal window.
    - The smoothed raster is reprojected and resampled to match a 10 m digital
      elevation model (DEM), ensuring full compatibility with local-scale
      modelling.

 6. Compute Model Residuals for Downscaling:
    - Predicted broad-scale probabilities are extracted back to the PA points.
    - Residuals are computed as:
         residual = observed_presence − predicted_probability
    - These residuals identify areas where the broad-scale climate model
      under- or over-predicts birch presence and form the response variable for
      the local-scale downscaling model.

 Features:
 - Produces a robust national-scale suitability model based on climate and
   large-scale environmental gradients.
 - Uses spatially explicit cross-validation to avoid inflated accuracy.
 - Generates residuals that allow fine-scale drivers (terrain, microclimate,
   landform structure) to be incorporated via local modelling.

 Output:
 - broad_scale_rf_prob.tif — national-scale birch suitability raster.
 - Random forest model object
 - residuals — point-level residuals used as input for the local downscaling
   model.


```{r include = F}
# Assigning presence and absence

pa_climate <- terra::extract(env_stack, pa_data, xy = TRUE)
pa_climate$presence <- pa_data$presence
pa_climate <- na.omit(pa_climate)
pa_climate$presence <- factor(pa_climate$presence, levels = c(0,1))
pa_climate <- pa_climate %>% dplyr::select(-c(ID,x,y))


```

```{r include = F}
# fitting the model

set.seed(123)
rf_broad_model <- randomForest(
presence ~ .,
data = pa_climate,
ntree = 500,
mtry = floor(sqrt(ncol(pa_climate))),  # fewer variables per split
nodesize = 10,  
importance = TRUE
)

importance(rf_broad_model)
varImpPlot(rf_broad_model, type = 1)

saveRDS(rf_broad_model, file = "models/broad_scale_rf.rds")
```

### Cross validation

```{r}

# 1) Convert presence/absence points to sf and project to raster CRS
pa_sf <- st_as_sf(pa_data)  
pa_sf <- st_transform(pa_sf, st_crs(env_stack))

# 2) Spatial blocking
set.seed(123)
sb <- spatialBlock(
  speciesData = pa_sf,
  species = "presence",
  rasterLayer = env_stack[[1]],  # defines extent
  theRange = 50000,              # block size in meters
  k = 5,                         # number of folds
  selection = "random",
  iteration = 100,
  showBlocks = TRUE
)
pa_sf$foldID <- sb$foldID

# 3) Split into training/testing for fold 1
train <- pa_sf[pa_sf$foldID != 1, ]
test  <- pa_sf[pa_sf$foldID == 1, ]

# 4) Convert sf to SpatVector for terra::extract
train_vect <- vect(train)
test_vect  <- vect(test)

# 5) Extract environmental variables and bind presence column
train_env <- terra::extract(env_stack, train_vect, bind = TRUE)
test_env  <- terra::extract(env_stack, test_vect, bind = TRUE)

# 6) Convert to data.frame for dplyr manipulation
train_env <- as.data.frame(train_env)
test_env  <- as.data.frame(test_env)

# 7) Remove ID column if it exists, then remove rows with NA
cols_to_remove <- intersect(c("ID"), colnames(train_env))
if(length(cols_to_remove) > 0) train_env <- train_env %>% select(-all_of(cols_to_remove))
train_env <- na.omit(train_env)

cols_to_remove <- intersect(c("ID"), colnames(test_env))
if(length(cols_to_remove) > 0) test_env <- test_env %>% select(-all_of(cols_to_remove))
test_env <- na.omit(test_env)

# 8) Convert presence to factor
train_env$presence <- factor(train_env$presence, levels = c(0,1))
test_env$presence  <- factor(test_env$presence, levels = c(0,1))

# 9) Quick NA check
colSums(is.na(train_env))  # should be all 0
colSums(is.na(test_env))   # should be all 0

# 10) Fit Random Forest
rf_model <- randomForest(
  presence ~ .,
  data = pa_climate,
  ntree = 500,
  mtry = floor(sqrt(ncol(pa_climate))),  # fewer variables per split
  nodesize = 10,  
  importance = TRUE)

# 11) Predict on test set
pred <- predict(rf_model, newdata = test_env, type = "response")

# 12) Evaluate performance
confusionMatrix(pred, test_env$presence)

```

### Prediction

```{r}
# predicting over Iceland

broad_scale_rf_prob <- terra::predict(
  env_stack,             # climate predictors raster stack
  rf_broad_model,                # your trained RF model
  type = "prob",
  index = 2, # probability of presence class
  filename = "data/broad_scale_rf_prob_birch.tif",  # write output directly to file
  overwrite = TRUE)

```

### smooths edges

```{r}

# Broadscale model (already loaded)
broad_prob <- broad_scale_rf_prob

# Reproject to a metric CRS (EPSG:8088 – Iceland Lambert 2016)
broad_proj <- terra::project(broad_prob, "EPSG:8088")

# Fill small NA gaps using a local mean
broad_filled <- terra::focal(
  broad_proj,
  w = 3,
  fun = function(x, ...) {
    if (is.na(x[5])) mean(x, na.rm = TRUE) else x[5]
  },
  na.policy = "omit",
  fillvalue = NA
)

# Define a 5 km smoothing radius
radius_m <- 5000
w <- terra::focalMat(broad_filled, d = radius_m, type = "circle")

# Apply smoothing (mean filter)
broad_smooth <- terra::focal(broad_filled, w = w, fun = mean, na.rm = TRUE)

# Reproject smoothed raster back to WGS84 (or original CRS of broad_prob)
broad_smooth_aligned <- terra::project(broad_smooth, crs(broad_prob), method = "bilinear")

# Resample smoothed raster to exactly match broad_prob (extent & resolution)
broad_smooth_aligned <- terra::resample(broad_smooth_aligned, broad_prob, method = "bilinear")

# Cover edges: keep broad_prob, fill NAs with smoothed raster
broad_edges_added <- terra::cover(broad_prob, broad_smooth_aligned)


```


```{r}
dem <- rast("data/IslandsDEMv1.0_10x10m_isn2016_zmasl.tif")

broad_scale_rf_prob_10m <- terra::project(broad_edges_added, dem)

```


```{r}

plot(broad_scale_rf_prob_10m)

```

```{r}
# ---- Predict probability of presence at each point used in training ----
pa_climate$pred_prob <- predict(rf_broad_model, type = "prob")[, 2]  # column 2 = presence class prob

# ---- Compute residuals ----
# presence is coded as 0 (pseudo-absence) or 1 (presence)
# residual = observed - predicted
pa_climate$residual <- as.numeric(as.character(pa_climate$presence)) - pa_climate$pred_prob

```

```{r}

# Convert residuals to SpatVector with coordinates
resid_vect <- pa_data  # same geometry as presence/pseudoabsence points
resid_vect$residual <- pa_climate$residual


```

## GEE 

This code chunk outputs the points needed for GEE.


```{r eval = F}
# export points for future work with GEE - Multiple mechanisms of early plant community assembly with stochasticity driving the process

sampled_forest_cells <- project(pa_data[, FALSE], "EPSG:4326")
writeVector(sampled_forest_cells, "data/GEE/SDM_pa_points_birch.shp", overwrite = TRUE)

```

### Fine-scale SDM-residual modelling

Fine-Scale terrain and satellite embeddings Species Distribution Model (SDM) residual Predictor Construction
#

 Purpose:
 This code prepares high-resolution environmental predictors for fine-scale 
 species distribution modeling. The predictors include topographic variables 
 (DEM, slope, TWI) and solar radiation variables (mean growing season and 
 peak summer radiation). These predictors will later be combined with embeddings 
 and residuals to train fine-scale SDMs.

 Workflow:

 1. Topography
    - Compute slope from the DEM using terra::terrain().
    - Aggregate the DEM to 100 m resolution for use in radiation modeling.
    - Save DEM at 100 m for reproducibility and use in SAGA GIS functions.

 2. Solar Radiation (SAGA GIS: Potential Incoming Solar Radiation)
    - Define representative mid-month days for May–September.
    - Loop over each mid-month day to compute:
        * Total radiation
        * Direct radiation
        * Diffuse radiation
        * Duration of sunlight
    - Use SAGA's `rsaga.pisr2` function with hourly integration.
    - Aggregate all monthly rasters to compute:
        * Mean growing season radiation (May–September)
        * Peak summer radiation (June–July)
    - Upscale radiation rasters to match DEM resolution (10 m) for SDM.

 3. Topographic Wetness Index (TWI)
    - Compute slope and aspect using SAGA's "Slope, Aspect, Curvature" module.
    - Compute upslope contributing area (Specific Catchment Area, SCA) using SAGA hydrology tools.
    - Compute TWI using slope and SCA to capture soil moisture potential and hydrological context.
    - Resulting TWI raster is aligned with DEM and other predictors.

 4. Outputs
    - High-resolution predictor rasters (10 m):
        * DEM_10m (original)
        * Slope (radians)
        * TWI
        * Mean growing season radiation (May–Sep)
        * Peak summer radiation (June–July)
    - Saved both as SAGA grids for intermediate steps and as GeoTIFFs for SDM use.

 Notes:
 - Solar radiation computations assume latitude ~65° (Iceland) and hourly steps.
 - Upscaling/downscaling ensures all predictor rasters are aligned for stacking.
 - These predictors provide both energy (radiation) and topographic context
   necessary for fine-scale ecological modeling.


## Elevation and slope

```{r}

# Compute slope and aspect
slope <- terrain(dem, v = "slope", unit = "radians")
dem_100m <- aggregate(dem, fact = 10, fun = mean)  # 100 m resolution
writeRaster(dem_100m, filename = "data/saga_output/DEM_100m.tif", overwrite = TRUE)

```

## Mean growing season and peak summer radiation 

```{r}

# set environment (adjust path if needed)
env <- rsaga.env(path = "C:/Users/jay.philip/Desktop/utilities/saga-9.10.0_x64/saga-9.10.0_x64")   # auto-detects SAGA; or rsaga.env(path = "C:/path/to/saga")
outdir <- "data/saga_output/"

# -----------------------
# 3. Define representative mid-month days
# -----------------------
# List of months and representative day of month
mid_month_days <- data.frame(
  month = 5:9,         # May–September
  day   = c(15, 15, 15, 15, 15)
)

# Store paths to Total radiation rasters
total_rasters <- c()
jun_jul_rasters <- c()

# -----------------------
# 4. Loop over each mid-month day
# -----------------------
for(i in 1:nrow(mid_month_days)) {
  m <- mid_month_days$month[i]
  d <- mid_month_days$day[i]
  
  out_total <- file.path(outdir, paste0("Total_", sprintf("%02d", m), ".tif"))
  out_direct <- file.path(outdir, paste0("Direct_", sprintf("%02d", m), ".tif"))
  out_diffuse <- file.path(outdir, paste0("Diffuse_", sprintf("%02d", m), ".tif"))
  out_duration <- file.path(outdir, paste0("Duration_", sprintf("%02d", m), ".tif"))
  
  cat("Computing solar radiation for month", m, "day", d, "\n")
  
  rsaga.pisr2(
    in.dem         = "data/saga_output/DEM_100m.tif",
    out.total.grid = out_total,
    out.direct.grid = out_direct,
    out.diffuse.grid = out_diffuse,
    out.duration    = out_duration,
    latitude       = 65,               # Iceland
    unit           = "kWh/m2",
    time.range     = c(0, 24),
    time.step      = 1,                # hourly integration
    start.date     = list(day = d, month = m, year = 2024),
    day.step       = 1,
    env            = env
  )
  
  # Collect raster paths for later averaging
  total_rasters <- c(total_rasters, out_total)
  if(m %in% 6:7) {  # June–July for peak summer
    jun_jul_rasters <- c(jun_jul_rasters, out_total)
  }
}

# -----------------------
# 5. Aggregate Total radiation rasters
# -----------------------

# Stack all Total rasters into a single SpatRaster for growing season
rast_stack <- do.call(c, lapply(total_rasters, rast))
growing_season_mean <- app(rast_stack, mean, na.rm = TRUE)
writeRaster(growing_season_mean, file.path(outdir, "Mean_GrowingSeason.tif"), overwrite = TRUE)

# Stack June–July rasters for peak summer
jun_jul_stack <- do.call(c, lapply(jun_jul_rasters, rast))
peak_summer_mean <- app(jun_jul_stack, mean, na.rm = TRUE)
writeRaster(peak_summer_mean, file.path(outdir, "Mean_JuneJuly.tif"), overwrite = TRUE)

cat("Finished computing growing season and peak summer radiation rasters.\n")


```

```{r}

# -----------------------
# 1. Read the averaged rasters
# -----------------------
growing_season <- rast(file.path(outdir, "Mean_GrowingSeason.tif"))
peak_summer   <- rast(file.path(outdir, "Mean_JuneJuly.tif"))

# -----------------------
# 2. Plot side by side
# -----------------------
par(mfrow = c(1, 2))  # two plots in one row

plot(growing_season, main = "Mean Growing Season Radiation\n(May–Sep, kWh/m²/day)")
plot(peak_summer, main = "Peak Summer Radiation\n(June–July, kWh/m²/day)")

par(mfrow = c(1,1))  # reset plotting layout


```

```{r}

# Upscale mean rasters to 10m
growing_season_10m <- resample(growing_season, dem, method = "bilinear")
peak_summer_10m   <- resample(peak_summer, dem, method = "bilinear")

# Write out upscaled rasters
writeRaster(growing_season_10m, file.path(outdir, "Mean_GrowingSeason_10m.tif"), overwrite = TRUE)
writeRaster(peak_summer_10m,   file.path(outdir, "Mean_JuneJuly_10m.tif"), overwrite = TRUE)

```

## Topographic wetness index

```{r eval=F}

# -----------------------
# SDM Topographic Predictor Preparation
# Computes: slope, aspect, upslope area (SCA), TWI
# -----------------------

library(RSAGA)
library(terra)

# -----------------------
# 1. Paths and environment
# -----------------------
dem_path <- "data/IslandsDEMv1.0_10x10m_isn2016_zmasl.tif"
dir.create(outdir, showWarnings = FALSE)

# -----------------------
# 2. Slope & Aspect
# -----------------------
slope_out  <- file.path(outdir, "Slope.sgrd")
aspect_out <- file.path(outdir, "Aspect.sgrd")

rsaga.geoprocessor(
  lib    = "ta_morphometry",
  module = "Slope, Aspect, Curvature",
  param  = list(
    ELEVATION = dem_path,
    SLOPE     = slope_out,
    ASPECT    = aspect_out
  ),
  env = env,
  show.output.on.console = TRUE
)

# -----------------------
# 3. Upslope Area (Specific Catchment Area)
# -----------------------
sca_out <- file.path(outdir, "SCA.sgrd")

rsaga.geoprocessor(
  lib    = "ta_hydrology",
  module = "Flow Width and Specific Catchment Area",
  param  = list(
    DEM        = dem_path,
    WIDTH      = NULL,        # optional
    TCA        = NULL,        # optional
    SCA        = sca_out,     # output specific catchment area
    COORD_UNIT = 0,
    METHOD     = 0
  ),
  env = env,
  show.output.on.console = TRUE
)

# -----------------------
# 4. TWI Calculation
# -----------------------
twi_out <- file.path(outdir, "TWI.sgrd")

rsaga.geoprocessor(
  lib    = "ta_hydrology",
  module = "Topographic Wetness Index",
  param  = list(
    SLOPE  = slope_out,
    AREA   = sca_out,   # use SCA for TWI
    TRANS  = 0,
    TWI    = twi_out,
    CONV   = 1,
    METHOD = 0
  ),
  env = env,
  show.output.on.console = TRUE
)

```

```{r}

twi <- rast("data/saga_output/TWI.tif")

```

### Modelling residuals using XGBoost

Purpose:
This chunk trains a fine-scale model to predict residuals from a broad-scale
model using both raster-derived environmental predictors and spatial embeddings.
The goal is to capture local variation not explained by the broad-scale model.

Workflow:
 1. Raster predictors:
    - Stack DEM, slope, peak summer temperature, and TWI.
    - Extract raster values at the locations of residual observations.
    - Remove any incomplete data points.

 2. Embeddings:
    - Load GEE-derived embeddings from CSV.
    - Clean and convert to sf points.
    - Ensure CRS matches residual points for spatial operations.

 3. Spatial join:
    - Assign nearest embedding to each residual observation.
    - Combine raster predictor values with embeddings.

 4. Prepare XGBoost data:
    - Extract predictor matrix and target residual vector.
    - Create DMatrix for XGBoost training.

 5. Model training:
    - Train a gradient boosting model (XGBoost) to predict residuals.
    - Parameters tuned for regression with moderate depth, learning rate,
      and subsampling for robustness.

 6. Feature importance:
    - Evaluate predictor contributions using XGBoost's importance metrics.

 Output:
 - xgb_fine_model_res: trained XGBoost residual model
 - importance_matrix / plot: feature importance for interpretation

 Notes:
 - Residual modeling focuses on explaining fine-scale variation not captured
   by broad-scale predictions.
 - Embeddings encode complex spatial/contextual patterns and improve prediction
   of residuals at local scales.

```{r}

# ===========================
# 1. Raster predictors
# ===========================
env_stack <- c(dem, slope, peak_summer_10m, twi)
names(env_stack) <- c("DEM", "Slope", "peak_summer", "twi")

# Extract raster values at residual points
resid_env <- terra::extract(env_stack, resid_vect, xy = TRUE)

# Add residuals from resid_vect
resid_env$residual <- resid_vect$residual

# Remove incomplete rows
resid_env <- resid_env %>% dplyr::filter(dplyr::if_all(dplyr::everything(), ~ !is.na(.)))

# ===========================
# 2. Embeddings
# ===========================
embeddings <- read.csv("data/GEE/birch_embeddings_2023_10m_avg.csv") %>%
  dplyr::rename(x = longitude, y = latitude) %>%
  dplyr::select(-dplyr::any_of(c("FID", "system.index","year", ".geo")))

# Convert to sf
embeddings_sf <- sf::st_as_sf(embeddings, coords = c("x", "y"), crs = 4326) # assume embeddings in WGS84

# Convert residuals to sf
resid_sf <- sf::st_as_sf(resid_vect)

# ===========================
# 3. CRS alignment
# ===========================
# Transform embeddings to match residuals CRS
embeddings_sf <- sf::st_transform(embeddings_sf, sf::st_crs(resid_sf))

# ===========================
# 4. Spatial join (nearest neighbor)
# ===========================
resid_with_embed_sf <- sf::st_join(resid_sf, embeddings_sf, join = sf::st_nearest_feature)

# Drop geometry
resid_with_embed_df <- sf::st_drop_geometry(resid_with_embed_sf)

# ===========================
# 5. Combine raster predictors + embeddings
# ===========================
resid_env_sf <- sf::st_as_sf(resid_env, coords = c("x", "y"), crs = sf::st_crs(resid_sf))

# Join raster predictors to embeddings
full_sf <- sf::st_join(resid_env_sf, embeddings_sf, join = sf::st_nearest_feature)

# Drop geometry to get final data frame
full_df <- sf::st_drop_geometry(full_sf)

# Verify residual column exists
if(!"residual" %in% colnames(full_df)) stop("Column 'residual' not found in full_df")

# ===========================
# 6. Prepare for XGBoost
# ===========================
resid_mat <- full_df %>%
  dplyr::select(-dplyr::any_of(c("residual","ID"))) %>%
  as.matrix()

resid_target <- full_df$residual

dtrain <- xgb.DMatrix(data = resid_mat, label = resid_target)

# ===========================
# 7. Train model
# ===========================
params <- list(
  objective = "reg:squarederror",
  eta = 0.1,
  max_depth = 4,
  subsample = 0.7,
  colsample_bytree = 0.7
)

set.seed(123)
xgb_fine_model_res <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 500,
  verbose = 1
)

# ===========================
# 8. Feature importance
# ===========================
importance_matrix <- xgb.importance(model = xgb_fine_model_res)
xgb.plot.importance(importance_matrix)


```

## Export AOI for GEE

### DEMO area only

```{r eval = F}

# Load regions
regions <- vect("data/boundaries/is_50v_mork_epsg_8088.gpkg/is_50v_mork_epsg_8088.gpkg", 
                layer = "mork_logregluumdaemi")

# Reproject to DEM CRS
regions_proj <- project(regions, crs(dem))

# Subset the region you want (2) reyjkanes
rykb_subset <- regions_proj[regions_proj$objectid == 3,]

# Convert to sf polygon (for GEE)
rykb_sf <- st_as_sf(rykb_subset)

# Split MULTIPOLYGON into individual POLYGON rows------
rykb_single_polys <- st_cast(rykb_sf, "POLYGON")
rykb_single_polys <- rykb_single_polys %>%
  mutate(area_m2 = st_area(geometry))

rykb_ordered <- rykb_single_polys %>%
  arrange(desc(area_m2))

second_largest_poly <- rykb_ordered[2, ]  # 2nd row after sorting
second_largest_poly <- second_largest_poly %>%
  select(-area_m2)

### combine it with reyjkanes

rykb_subset <- regions_proj[regions_proj$objectid == 2,]
rykb_sf <- st_as_sf(rykb_subset)

#merge 
merged_sf <- rbind(rykb_sf, second_largest_poly)
rykb_subset <- st_union(merged_sf)


# Save as GeoJSON for GEE upload
# st_write(rykb_subset, "data/GEE/complete_reykjanes_penisula.shp", delete_dsn = TRUE, overwrite = T)

```

## Read in Embeddings and create fine-scale topographic-embeddings stack -----DEMO AREA

```{r}

library(terra)

# Set directory
raster_dir <- "data/GEE/reyk_penisula_tiles"

# List all TIFF files
tif_files <- list.files(raster_dir, pattern = "\\.tif$", full.names = TRUE)

# Read all rasters into a list
rasters_list <- lapply(tif_files, rast)

# Merge tiles together (works even if extents don't match)
merged_raster <- do.call(merge, rasters_list)
env_stack <- crop(env_stack, rykb_subset)
# env_stack <- mask(env_stack, rykb_subset)

# Align CRS of merged_raster to env_stack
merged_raster_proj <- project(merged_raster, crs(env_stack))

# Resample merged_raster to match env_stack resolution and extent
merged_raster_resampled <- resample(merged_raster_proj, env_stack, method = "bilinear") 
# "bilinear" is good for continuous data; use "near" for categorical

# Crop to exact env_stack extent (optional, just to be safe)
merged_raster_crop <- crop(merged_raster_resampled, ext(env_stack))

# Stack with existing env_stack
full_stack <- c(env_stack, merged_raster_crop)


```

## Predict over AOI

```{r}
library(terra)

# Extract predictor values from raster for Reykjanes
pred_mat <- terra::values(full_stack, mat = TRUE)  # matrix of raster values

# Remove rows with NA (XGBoost cannot handle NAs)
na_rows <- apply(pred_mat, 1, function(x) any(is.na(x)))
pred_mat_clean <- pred_mat[!na_rows, , drop = FALSE]

# Keep track of which cells are valid
valid_cells <- which(!na_rows)

# Predict residuals
resid_pred <- predict(
  xgb_fine_model_res,
  newdata = pred_mat_clean
)

# Create an empty raster to store predictions
resid_raster <- full_stack[[1]]  # template
terra::values(resid_raster) <- NA
terra::values(resid_raster)[valid_cells] <- resid_pred

# Plot
plot(resid_raster, main = "Fine-scale Residual Prediction (Gradient Boosting)")
terra::writeRaster(resid_raster, "data/residual_plot_embedded.tif", overwrite = TRUE)

```

## Resample broad-scale model

```{r}

# Ensure CRS match
broad_scale_rf_prob_10m <- project(broad_scale_rf_prob_10m, crs(rykb_subset))

# Crop to the Reykjanes bounding box
broad_crop <- crop(broad_scale_rf_prob_10m, rykb_subset)
# broad_crop <- mask(broad_crop, rykb_subset)

plot(broad_crop, main = "Broad-scale probability (Reykjanes subset)")

```

```{r}

writeRaster(broad_crop, "data/broad_scale_model_cropped.tif", overwrite=TRUE)

```

## Combine residuals to broad-scale probabilities and construct final predictions for demo area

```{r}

# Align grid, extent, and resolution
fine_resid_pred <- project(resid_raster, crs(broad_crop))
fine_resid_pred <- resample(resid_raster, broad_crop, method = "bilinear")

# Combine
multiscale_reykjanes <- broad_crop + fine_resid_pred

# Clamp to valid probability range
multiscale_reykjanes <- clamp(multiscale_reykjanes, 0, 1)

# Plot
plot(multiscale_reykjanes, main = "Multiscale Betula SDM (Reykjanes, 10 m)")

# Save output
writeRaster(multiscale_reykjanes,
            "data/multiscale_reykjanes_10m_embeddings_predictions.tif",
            overwrite = TRUE)

```

## Fit fine-scale predictions across Iceland

Overview:
This pipeline processes large raster embedding datasets (e.g., GEE embeddings)
together with environmental predictor rasters to produce fine-scale predicted
surfaces for multiple regions of Iceland. It is designed to handle regions
that are too large to fit entirely in memory by splitting them into smaller
manageable chunks.

Workflow:
1. Detect all region folders containing embedding tiles (e.g., "westfjords_tiles").
2. Merge all embedding raster tiles within a region into a single raster.
3. Split (chunk) the merged embedding raster adaptively:
    - The number of chunks per dimension is determined based on a user-defined
      maximum chunk size (`max_cells`) to keep memory usage manageable.
    - Each chunk is saved as a temporary raster (embeddings only).
4. For each chunk:
    - Crop and resample the environmental stack to match the chunk extent and CRS.
    - Combine the environmental predictors with the embedding chunk to form
      the full predictor stack.
    - Extract values and skip the chunk if it contains only NA (e.g., ocean areas).
    - Predict using the XGBoost model and write a chunk-level prediction raster.
5. Mosaic all predicted chunks for a region into a single **final prediction raster**.
    - This raster covers the entire region, preserving spatial alignment.
    - Areas with no data (e.g., ocean) remain NA.
6. Repeat for all region folders.

Features:
 - Adaptive chunking: automatically adjusts chunk size to balance memory usage
   and processing speed.
 - Skips empty chunks automatically to avoid XGBoost warnings on ocean tiles.
 - Produces one final seamless prediction raster per region.
 - Intermediate chunk rasters can be optionally cleaned up to save disk space.

Output:
 predictions/
   ├── <region_folder>/
   │    ├── chunks/           <- embedding chunks (optional intermediate)
   │    ├── pred_chunks/      <- predicted chunk rasters (optional intermediate)
   │    └── final_prediction.tif  <- final mosaic of predictions for the region

 Notes:
 - `max_cells` controls the maximum number of columns/rows per chunk and
   therefore indirectly controls memory usage per chunk.
 - The final mosaic is independent of chunk size; more chunks only affect
   memory management and I/O, not prediction accuracy.

```{r}

# Load regions
regions <- vect("data/boundaries/is_50v_mork_epsg_8088.gpkg/is_50v_mork_epsg_8088.gpkg", 
                layer = "mork_logregluumdaemi")

# Reproject to DEM CRS
regions_proj <- project(regions, crs(dem))

# Subset the region you want (2) reyjkanes
rykb_subset <- regions_proj[regions_proj$objectid == 3,]
rykb_sf <- st_as_sf(rykb_subset)
plot(rykb_sf)

# Convert to sf polygon (for GEE)
st_write(rykb_sf, "data/GEE/ryk.shp", delete_dsn = TRUE, overwrite = T)

```

```{r}

##############################################################
## 1. Adaptive chunking function
##############################################################
chunk_raster_adaptive <- function(r, max_cells = 1000000, out_dir) {
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  
  # Compute side length in cells for square chunks
  side_cells <- ceiling(sqrt(max_cells))
  
  # Compute number of chunks along x and y
  n_chunks_x <- ceiling(ncol(r) / side_cells)
  n_chunks_y <- ceiling(nrow(r) / side_cells)
  
  ex <- ext(r)
  xstep <- (ex$xmax - ex$xmin) / n_chunks_x
  ystep <- (ex$ymax - ex$ymin) / n_chunks_y
  
  out_files <- c()
  
  for (i in 0:(n_chunks_x-1)) {
    for (j in 0:(n_chunks_y-1)) {
      
      e <- ext(
        ex$xmin + i * xstep,
        ex$xmin + (i+1) * xstep,
        ex$ymin + j * ystep,
        ex$ymin + (j+1) * ystep
      )
      
      ch <- crop(r, e)
      outfile <- file.path(out_dir, paste0("chunk_", i, "_", j, ".tif"))
      writeRaster(ch, outfile, overwrite = TRUE)
      out_files <- c(out_files, outfile)
    }
  }
  
  return(out_files)
}

##############################################################
## 2. Build full predictor stack for a chunk
##############################################################
stack_for_chunk <- function(env_stack, chunk_emb) {
  chunk_emb <- project(chunk_emb, env_stack)
  chunk_emb <- resample(chunk_emb, env_stack)
  env_crop <- crop(env_stack, chunk_emb)
  c(env_crop, chunk_emb)
}

##############################################################
## 3. Predict for one chunk (skip empty chunks)
##############################################################
predict_chunk <- function(full_stack, model, outfile) {
  vals <- terra::values(full_stack, mat = TRUE)
  valid <- complete.cases(vals)
  vals_clean <- vals[valid,, drop = FALSE]
  
  if (nrow(vals_clean) == 0) {
    message("Skipping empty chunk: ", outfile)
    return(NULL)
  }
  
  preds <- predict(model, vals_clean)
  
  out_r <- full_stack[[1]]
  values(out_r) <- NA
  values(out_r)[valid] <- preds
  
  writeRaster(out_r, outfile, overwrite = TRUE)
  return(outfile)
}

##############################################################
## 4. Process one area folder
##############################################################
process_area_adaptive <- function(area_dir, env_stack, model, max_cells = 1000000, outdir_base="predictions") {
  
  message("Processing area: ", area_dir)
  
  # Merge embeddings in area folder
  tifs <- list.files(area_dir, "\\.tif$", full.names = TRUE)
  rs <- lapply(tifs, rast)
  area_emb <- do.call(merge, rs)
  
  # Adaptive chunking
  chunk_dir <- file.path(outdir_base, basename(area_dir), "chunks")
  chunk_files <- chunk_raster_adaptive(area_emb, max_cells = max_cells, out_dir = chunk_dir)
  
  # Predict per chunk
  pred_dir <- file.path(outdir_base, basename(area_dir), "pred_chunks")
  dir.create(pred_dir, recursive = TRUE, showWarnings = FALSE)
  
  pred_files <- c()
  
  for (cf in chunk_files) {
    message("  predicting chunk: ", cf)
    
    chunk_emb <- rast(cf)
    full_stack <- stack_for_chunk(env_stack, chunk_emb)
    
    out_file <- file.path(pred_dir, paste0("pred_", basename(cf)))
    pf <- predict_chunk(full_stack, model, out_file)
    
    if (!is.null(pf)) pred_files <- c(pred_files, pf)
  }
  
  # Mosaic predictions back together
  message("  mosaicing predicted chunks...")
  if (length(pred_files) == 0) {
    message("  No valid chunks to mosaic for area: ", area_dir)
    return(NULL)
  }
  
  r_list <- lapply(pred_files, rast)
  final <- do.call(mosaic, r_list)
  
  final_out <- file.path(outdir_base, basename(area_dir), "final_prediction.tif")
  writeRaster(final, final_out, overwrite = TRUE)
  
  return(final_out)
}

##############################################################
## 5. Run the pipeline for all area folders
##############################################################
run_all_areas_adaptive <- function(base_dir, env_stack, model, max_cells = 1000000, outdir="predictions") {
  
  # Detect area folders containing "tiles"
  areas <- list.dirs(base_dir, recursive = FALSE)
  areas <- areas[grepl("tiles", basename(areas), ignore.case = TRUE)]
  
  results <- lapply(areas, function(a) {
    process_area_adaptive(a, env_stack, model, max_cells, outdir)
  })
  
  return(results)
}

##############################################################
## 6. Run it
##############################################################
final_outputs <- run_all_areas_adaptive(
  base_dir = "data/GEE",
  env_stack = env_stack,
  model = xgb_fine_model_res,
  max_cells = 1000000,          # Adjust to control chunk size in memory
  outdir = "predictions"
)

```


```{r}

library(terra)
library(xgboost)
library(parallel)

##############################################################
## 1. FAST adaptive chunk creation (window-based, no crop)
##############################################################
chunk_raster_adaptive <- function(r, max_cells = 1e6, out_dir) {
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  
  side_cells <- ceiling(sqrt(max_cells))
  n_chunks_x <- ceiling(ncol(r) / side_cells)
  n_chunks_y <- ceiling(nrow(r) / side_cells)
  
  out_files <- vector("list", n_chunks_x * n_chunks_y)
  id <- 1
  
  for (i in seq_len(n_chunks_x)) {
    for (j in seq_len(n_chunks_y)) {

      row_start <- (j - 1) * side_cells + 1
      col_start <- (i - 1) * side_cells + 1

      # window() = correct pixel-based cropping
      win <- terra::window(
        r,
        row = row_start,
        nrows = side_cells,
        col = col_start,
        ncols = side_cells
      )

      outfile <- file.path(out_dir, paste0("chunk_", i-1, "_", j-1, ".tif"))
      writeRaster(win, outfile, overwrite = TRUE)
      out_files[[id]] <- outfile
      id <- id + 1
    }
  }
  
  out_files
}

##############################################################
## 2. Build predictor stack
##############################################################
stack_for_chunk <- function(env_stack, chunk_emb) {

  if (!identical(crs(env_stack), crs(chunk_emb))) {
    message("  Reprojecting chunk...")
    chunk_emb <- project(chunk_emb, env_stack)
  }

  chunk_emb <- resample(chunk_emb, env_stack)
  env_crop <- crop(env_stack, ext(chunk_emb))

  c(env_crop, chunk_emb)
}

##############################################################
## 3. Predict with batching 
##############################################################
predict_chunk <- function(full_stack, model, outfile, batch = 50000) {

  vals <- terra::values(full_stack, mat = TRUE)
  valid <- complete.cases(vals)

  if (!any(valid)) {
    message("Skipping empty chunk: ", outfile)
    return(NULL)
  }

  vals_clean <- vals[valid,, drop = FALSE]
  n <- nrow(vals_clean)
  preds <- numeric(n)

  starts <- seq(1, n, by = batch)
  for (s in starts) {
    e <- min(s + batch - 1, n)
    preds[s:e] <- predict(model, vals_clean[s:e, , drop = FALSE])
  }

  out_r <- full_stack[[1]]
  values(out_r) <- NA
  values(out_r)[valid] <- preds

  writeRaster(out_r, outfile, overwrite = TRUE)
  outfile
}

##############################################################
## 4. Process an area (reproject embedding only once)
##############################################################
process_area_adaptive <- function(area_dir, env_stack, model, 
                                  max_cells = 1e6, outdir_base="predictions",
                                  cores = 1) {

  message("Processing area: ", area_dir)

  # Merge embedding tiles
  tifs <- list.files(area_dir, "\\.tif$", full.names = TRUE)
  rs <- lapply(tifs, rast)
  area_emb <- do.call(merge, rs)

  # Reproject whole embedding raster once
  if (!identical(crs(env_stack), crs(area_emb))) {
    message("Reprojecting full embedding raster...")
    area_emb <- project(area_emb, env_stack)
  }

  # Create chunks
  chunk_dir <- file.path(outdir_base, basename(area_dir), "chunks")
  chunk_files <- chunk_raster_adaptive(area_emb, max_cells, chunk_dir)

  pred_dir <- file.path(outdir_base, basename(area_dir), "pred_chunks")
  dir.create(pred_dir, recursive = TRUE, showWarnings = FALSE)

  ############################################################
  ## Parallel or sequential prediction
  ############################################################
  predict_one <- function(cf) {
    message("  predicting chunk: ", cf)
    chunk_emb <- rast(cf)
    full_stack <- stack_for_chunk(env_stack, chunk_emb)
    out_file <- file.path(pred_dir, paste0("pred_", basename(cf)))
    predict_chunk(full_stack, model, out_file)
  }

  if (cores > 1) {
    message("Running in parallel on ", cores, " cores...")
    pred_files <- mclapply(chunk_files, predict_one, mc.cores = cores)
  } else {
    pred_files <- lapply(chunk_files, predict_one)
  }

  pred_files <- unlist(pred_files)

  ############################################################
  ## Mosaic chunks
  ############################################################
  message("  mosaicing predicted chunks...")
  r_list <- lapply(pred_files, rast)
  final <- do.call(mosaic, r_list)

  final_out <- file.path(outdir_base, basename(area_dir), "final_prediction.tif")
  writeRaster(final, final_out, overwrite = TRUE)

  final_out
}

##############################################################
## 5. Run all areas
##############################################################
run_all_areas_adaptive <- function(base_dir, env_stack, model, 
                                   max_cells = 1e6, outdir="predictions",
                                   cores = 1) {
  
  areas <- list.dirs(base_dir, recursive = FALSE)
  areas <- areas[grepl("tiles", basename(areas), ignore.case = TRUE)]
  
  lapply(areas, function(a) {
    process_area_adaptive(a, env_stack, model, max_cells, outdir, cores)
  })
}

##############################################################
## 6. Run it
##############################################################
final_outputs <- run_all_areas_adaptive(
  base_dir = "data/GEE",
  env_stack = env_stack,
  model = xgb_fine_model_res,
  max_cells = 1e6,
  outdir = "predictions",
  cores = 6        # <-- parallel threads
)


```